\chapter{User Personae}\label{append_personae}

\section{Primary Persona}
User: Steve Smith, Network Administrator. \\

Background:  Steve graduated from Victoria university with a BE in networking, Since graduating, he has been working for HB (A large software consultancy firm) as a network administrator. Computers and networks form his day to day world, and he has developed significant experience and expertise. 

Key Goals: Steve works hard to ensure that HB networks are kept secure, he is interested in any new tools which may improve his ability to understand activity on the systems he is responsible for. 

Usage scenarios:
\begin{itemize}
\item{Steve has been asked by his manager to investigate missing files from a network share. This has lead Steve to suspect that Sam (an employee of HB) has been misusing HB networks. Steve wants to confirm when Sam has been logging in, and where from, as part of tracking all of Sam's activity on HB networks.}
\item{Steve has been asked to investigate the source of unusually high traffic on the SSH ports for one of the HB servers. Steve wants to know how much traffic has been coming from each source, and if the traffic is malicious what accounts (if any) have been compromised}
\end{itemize}


\section{Secondary Persona}

User: Diana Wozniak, Network Manager \\

Background:  Diana has been employed at HB for many years, and now manages the IT security department including Steve. Her day to day job consists primarily of administration and management of staff, though she still likes to get involved in the actual security work when she can. 

Key Goals: Diana wants to stay current on tools used for security work, as this is her primary interest, though her job has taken her into a more managerial and administrative role. 

Usage scenario:
Steve has found traces of activity in logs which he is not sure is authorised. He has asked Diana for a second opinion, and possibly followup if required. Diana is seeking to understand what attracted Steve's attention, and why this activity could be harmful. 

\section{Anti-persona}
User: Adrian ``Agent\_zero'' Anderson, Hacker \\

Background: Adrian is a freelance hacker, working with criminal syndicates to assist with bypassing computer security. In his day job, he works as a developer for HB. Computers and security systems are a day to day part of his life, and he has developed considerable expertise with them. 

Key Goals: Adrian works as a hacker in order to earn extra money, He feels that HB are not paying him enough for his work, and enjoys the challenge of bypassing security systems.

Usage scenario: Adrian has been contracted by a syndicate to break into Grey Goo international,  a midsized manufacturing corporation. The breakin was requested to steal plans for Grey Goo's latest product. Adrian's purpose in using the tool is to gather more information about the network structure at Grey Goo, and harvest usernames for later attack.